\begin{thebibliography}{99}

\bibitem{BZ1}
Belousouv BP.
[A periodic reaction and its mechanism (Russian)].
[Compilation of Abstracts on Radiation Medicine], \textbf{147}:145 (1959)

\bibitem{Chickarmane}
Chickarmane V, Peterson C.
\textit{A Computational Model for Understanding Stem Cell, Trophectoderm and Endoderm Lineage Determination.}
\textbf{PLoS ONE 3}(10):e3478. \url{doi:10.1371/journal.pone.0003478}

\bibitem{Cornish-Bowden}
Cornish-Bowden A.
\textit{An automatic method for deriving steady-state rate equations}
\textbf{Biochem J. 165}:55-59 (1977)

\bibitem{Repressilator}
Elowitz MB, Leibler S.
\textit{A synthetic oscillatory network of transcriptional regulators.}
\textbf{Nature. 403}:335-338 (2000).

\bibitem{Oregonator1}
Field RJ, Koros RM, Noyes RM 
\textit{Oscillations in Chemical Systems II. Thorough analysis of temporal oscillations in the Ce-BrO3- - malonic acid system.}
\textbf{J. Am. Chem. Soc.}:94(25), 8649-64 (1972).

\bibitem{Oregonator2}
Field RJ, Noyes RM
\textit{Oscillations in Chemical Systems IV. Limit cycle behavior in a model of a real chemical reaction.}
\textbf{J. Chem. Phys. 60}: 1877-84 (1974).

\bibitem{Goldbeter}
Goldbeter A.
\textit{A minimal cascade model for the mitotic oscillator involving cyclin and cdc2 kinase.}
\textbf{Proc. Natl. Acad. Sci. USA 88}:9107-1101 (1991).

\bibitem{NFKB}
Hoffmann A, Levchenko A, Scott ML, Baltimore D. 
\textit{THe IkappaB-NF-kappaB signaling module:temporal control and selective gene activation.}
\textbf{Science. 298}: 1241 (2002).

\bibitem{SBML2}
Hucka M, Hoops S, Keating SM, Le Nov\`ere N, Sahle S, Wilkinson DJ.
\textit{Systems Biology Markup Language (SBML) Level 2: Structures and Facilities for Model Definitions. }
\textbf{Nature Precedings} (doi:10.1038/npre.2008.2715.1) \url{http://precedings.nature.com/documents/2715/version/1}

\bibitem{King-Altman}
King EL, Altman C.
\textit{A Schematic Method of Deriving the Rate Laws for Enzyme-Catalyzed Reactions.}
\textbf{J. Phys. Chem. 60}: 1375-1378 (1956).

\bibitem{ecoli}
Ko CL, Wang F-S, Chao Y-P, Chen T-W.
\textit{S-system approach to modeling recombinant Escherichia coli growth by
hybrid differential evolution with data collocation}
\textbf{Biochem. Eng. J.}, 28:10-16 (2006).


\bibitem{MWC}
Monod J, Wyman J, Changeux JP. 
\textit{On the Nature of Allosteric Transitions: A Plausible Model.}
\textbf{J. Mol. Bio. 12}:88-118 (1965). 

\bibitem{GMWC}
Najdi TS, Yang CR, Shapiro BE, Hatfield W, Mjolsness E.
\textit{Application of a Generalized MWC Model for the Mathematical Simulation of Metabolic Pathways Regulated by Allosteric Enzymes.}
\textbf{J. Bioinf. Comp. Bio. 4}(2): 335-356 (2006). 

\bibitem{SSYS1}
Savageau MA.
\textit{Biochemical systems analysis. II. The steady-state solutions for an n-pool system using a power-law approximation.}
\textbf{J Theor Biol 25}:370-379 (1969).

\bibitem{SSYS2}
Savageau MA. 
\textit{Biochemical systems analysis. 3. Dynamic solutions using a power-law approximation}. 
\textbf{J Theor Biol 26}:215-226 (1970).


\bibitem{Cellerator}
Shapiro BE, Levchenko A, Meyerowitz EM, Wold BJ, Mjolsness ED.
\textit{Cellerator: extendeing a computer algebra system to include biochemical arrows for signal transduction simulations}.
\textbf{Bioinformatics, 19}(5):677-8 (2003). (doi: 10.1093/bioinformatics/btg042)

\bibitem{ICSB2001}
Shapiro BE, Mjolsness ED.
\textit{Developmental Simulations with Cellerator}.
Second International Conference on Systems Biology (ICSB-2001), Pasadena, CA, USA (2001). 

\bibitem{kmech}
Yang CR, Shapiro BE, Mjolsness ED, Hatfield ED.
\textit{AN enzyme mechanism language for the mathematical modeling of metabolic pathways}.
\textbf{Bioinformatics, 21}(6):774-780 (2005) (doi: 10.1093/bioinformatics/bti068)

\bibitem{GY-Diss}
Yosiphon G.
\textbf{Stochastic Parameterized Grammars : Formalization, Inference and Modeling Applications}, Ph.D. Thesis, Univ. Cal., Irvine (2009). 

\bibitem{GY-MIT}
Yosiphon G, Mjolsness E.
\textit{Towards the Inference of Stochastic Biochemical Network and Parameterized Grammar Models}
in \textbf{Learning and Inference in Computational Systems Biology},
ed. N. Lawrence et. al. 
 MIT Press (2010).
 
 \bibitem{GY-INPREP}
Yosiphon G, Gokoffski KK, Lander AD, Calof AL, Mjolsness E. 
\textit{Dynamical Grammar Modeling of Growth and Lamination of the Olfactory Epithelium}.
\textbf{International Conference on Systems Biology (ICSB)} (2007).

\bibitem{BZ2}
Zhabotinsky AM.
[Periodic processes of malonic acid oxidation in a liquid phase. (Russian)]
[Biofizika],\textbf{9}:306-311 (1964).

\end{thebibliography}

