\copyright \ 2015 Bruce E Shapiro \\ \vspace{-8pt} \hrule
\begin{small}
This work is licensed under the Creative Commons Attribution-NonCommercial-NoDerivs 3.0 United States License. To view a copy of this license, visit \url{http://creativecommons.org/licenses/by-nc-nd/3.0/us/}  or send a letter to Creative Commons, 444 Castro Street, Suite 900, Mountain View, California, 94041, USA.\\
The following is a summary of the license.  

You are free:
\vspace{-12pt}
\begin{itemize}
\item to Share -- to copy, distribute and transmit the work
\end{itemize}
\vspace{-12pt}
Under the following conditions:
\vspace{-12pt}
\begin{itemize}
\setlength{\itemsep}{0pt}
\setlength{\parsep}{0pt}
\setlength{\topsep}{0pt}
\setlength{\parskip}{0pt}
\setlength{\partopsep}{0pt}
\item Attribution -- You must attribute the work in the manner specified by the author or licensor (but not in any way that suggests that they endorse you or your use of the work).
\item Noncommercial -- You may not use this work for commercial purposes.
\item No Derivative Works -- You may not alter, transform, or build upon this work.
\end{itemize}
\vspace{-12pt}
With the understanding that:
\vspace{-12pt}
\begin{itemize}
\setlength{\itemsep}{0pt}
\setlength{\parsep}{0pt}
\setlength{\topsep}{0pt}
\setlength{\parskip}{0pt}
\item Waiver -- Any of the above conditions can be waived if you get permission from the copyright holder.
\item Public Domain -- Where the work or any of its elements is in the public domain under applicable law, that status is in no way affected by the license.
\item Other Rights -- In no way are any of the following rights affected by the license:
\vspace{-4pt}
\begin{itemize}
\setlength{\itemsep}{0pt}
\setlength{\parsep}{0pt}
\setlength{\topsep}{0pt}
\setlength{\parskip}{0pt}
\item Your fair dealing or fair use rights, or other applicable copyright exceptions and limitations;
\item The author's moral rights;
\item Rights other persons may have either in the work itself or in how the work is used, such as publicity or privacy rights.
\end{itemize}
\end{itemize}
\vspace{-12pt} 
Notice -- For any reuse or distribution, you must make clear to others the license terms of this work. The best way to do this is with a link to the web page \url{http://creativecommons.org/licenses/by-nc-nd/3.0/us/}. \\

\hrule 
pycellerator is is free software: you can redistribute it and/or modify
    it under the terms of the GNU General Public License as published by
    the Free Software Foundation, either version 3 of the License, or
    (at your option) any later version.

    This program is distributed in the hope that it will be useful,
    but WITHOUT ANY WARRANTY; without even the implied warranty of
    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
    GNU General Public License for more details.
    
    A copy of the license is provide in Appendix \ref{chapter:GPL}. For more details visit \url{http://www.gnu.org/licenses/}. 
    
    \hrule
    
    The Python language is Copyright (c) 2001-2015 Python Software Foundation; All Rights Reserved.  The python open source licensed may be viewed at \url{https://www.python.org/download/releases/2.7/license/}.  


Anaconda Python is licensed and copyright by Continuum Analytics. The end user license agreement ma be viewed at \url{http://docs.continuum.io/anaconda/eula}. 

Enthought Canopy is licensed and copyright by Enthought, Inc. The end users license agreement may be viewed at \url{https://www.enthought.com/licenses/}. 

libSBML is copyright Caltech, EMBL, and JST and licensed under the LGPL. The end user license may be viewed at \url{http://sbml.org/Software/libSBML/LibSBML_License}. 

\end{small}
